Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

\begin{proof} \
  \TODO{this}
\end{proof}

\TODO{pull down to appendix}

@@

\begin{proof} \
  \TODO{this}
\end{proof}

\TODO{pull down to appendix?}

@@

\begin{proof} \
  \TODO{this}
\end{proof}

\TODO{pull down to appendix? This is just a trivial induction.}

@@

\begin{proof} \
  \TODO{this}
\end{proof}

@@

\begin{proof} \
  \TODO{this}
\end{proof}
\TODO{float proof to appendix}

@@

\begin{proof} \
  \TODO{this}
\end{proof}

\TODO{pull down to appendix? just trivial induction}

@@

Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of type \(A \to \ms{Heap} \to \ms{Prop}\). 
Taking as objects pairs \((A, \Phi)\), we can define the category \(\ms{Sep}\) to have morphisms

@@

is the usual partial correctness triple definition and the separating conjunction of heap-predicates is defined as usual to be 

@@

We can naturally define a tensor product on objects \((A, \Phi) \otimes (B, \Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking

@@

  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the predicate which holds only for the empty heap
  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B, \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is associative
  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} = \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\) is the unit of \(\ast\)
  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f \times C\), which are valid definitions by the frame rule

@@

  \item Affine objects to be precisely those for which \(\ms{drop} = \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
  \item Relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in this case those for which \(\Phi\) does not depend on the heap.
  \item Central morphisms to be those which do not print \textit{or} modify the heap. We will consider how to relax this condition later.

@@

  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a, \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1}, \ms{emp}))\)

@@

  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell, \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address at which the value passed in was stored to overcome a lack of dependency between the source and target objects.

@@

  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)

@@

We can clearly see that this is a refinement type system in the sence of \cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi) \mapsto A\) which is simply the identity on morphisms. Unfortunately, this means that our notion of equality is a little too stringent, since we need to consider equality on inputs which do not match our preconditions.

\TODO{example, + example set}

\TODO{restricted equality}

\TODO{sliding w/ properly restricted equality}

\TODO{\(\implies\) central = no printing, heap allowed: can make heap operations relevant/affine too!}

>>>>

@@

@@

@@

@@

@@

@@

Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
we can define the category \(\ms{Sep}\) to have morphisms

@@

is the usual partial correctness triple definition and the separating
conjunction of heap-predicates is defined as usual to be 

@@

We can naturally define a tensor product on objects \((A, \Phi) \otimes (B,
\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking

@@

  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the
  predicate which holds only for the empty heap
  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B,
  \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is
  associative
  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} =
  \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\)
  is the unit of \(\ast\)
  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f
  \times C\), which are valid definitions by the frame rule

@@

  \item Affine objects to be precisely those for which \(\ms{drop} =
  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for
  this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend
  on the heap.
  \item Relevant objects to be precisely those for which \(\ms{split} =
  \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in
  this case those for which \(\Phi\) does not depend on the heap.
  \item Central morphisms to be those which do not print, diverge, \textit{or} modify the
  heap. We will consider how to relax this condition later.

@@

  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a,
  \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1},
  \ms{emp}))\)

@@

  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell,
  \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell
  \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address
  at which the value passed in was stored to overcome a lack of dependency
  between the source and target objects.

@@

  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda
  \ell. \exists v. \ell \mapsto v))\)

@@

We can clearly see that this is a refinement type system in the sence of
\cite{mellies-ftrs}: in particular, there is a forgetful functor \((A, \Phi)
\mapsto A\) which is simply the identity on morphisms. Unfortunately, this means
that our notion of equality is a little too stringent, since we need to consider
equality on inputs which do not match our preconditions. Thankfully, we can fix
this problem quite easily by quotienting morphisms over the equivalence relation
  \forall f, g: \ms{Sep}((A, \Phi), (B, \Psi)).
  f \simeq g \iff \forall a, h. \Phi\;a\;h \implies f\;a\;h = g\;a\;h
yielding the new category \(\ms{SComp}\). In particular, this allows us to widen
our definition of central morphisms to include all morphisms which do not print
or diverge, including in particular \(\ms{set}, \ms{get}, \ms{alloc}\) and
\(\ms{free}\), which we can in fact mark relevant and affine as well, allowing
us, for example, to make use of intermediate allocations in an otherwise pure
computation.
